# Get related information

In this chapter we retrieve related information for taxa on the checklists (with a valid distribution).

## Read taxa

1. Read taxa from [`data/interim/taxa_with_verification.csv`](https://github.com/trias-project/unified-checklist/blob/master/data/interim/taxa_with_verification.csv).

```{r get_information-1}
taxa <- read_csv(here("data", "raw", "taxa.csv"))
```

2. Extract `taxonKeys` as a vector.

```{r get_information-2}
taxon_keys <-
  taxa %>%
  pull(taxonKey)
```

## Get distributions

This was already done in \@ref(filter-on-distribution), where only distributions of interest were kept.

## Get species profiles

1. Get species profiles for our taxa from GBIF.

```{r get_information-get_speciesprofiles_from_gbif, cache = TRUE}

progressr::with_progress({
  progress_bar <- progressor(steps = length(taxon_keys))
  speciesprofiles <-
    future_map_dfr(taxon_keys, function(x) {
      progress_bar()
      rgbif::name_usage(
        key = x,
        data = "speciesProfiles"
      )$data
  }, .options = furrr_options(seed = TRUE))
})
```

2. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/raw/speciesprofiles.csv).

```{r get_information-3}
write_csv(speciesprofiles, here("data", "raw", "speciesprofiles.csv"), na = "")
```

## Get descriptions

1. Get descriptions for our taxa from GBIF.

```{r get_information-get_descriptions_from_gbif, cache = TRUE}

progressr::with_progress({
  progress_bar <- progressor(steps = length(taxon_keys))
  descriptions <-
    future_map_dfr(taxon_keys, function(x) {
      progress_bar()
      rgbif::name_usage(
        key = x,
        data = "description"
      )$data
  }, .options = furrr_options(seed = TRUE))
})
```

2. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/raw/descriptions.csv).

```{r get_information-4}
write_csv(descriptions, here("data", "raw", "descriptions.csv"), na = "")
```
