# Verify taxa

In this chapter we verify taxonomic information returned by the [GBIF Backbone Taxonomy](https://doi.org/10.15468/39omei). We do this because we want to use the backbone information to unify taxa that can be considered the same within or across checklists. This is straightforward for (accepted) taxa with the same backbone key (`bb_key`), but requires verification for taxa the backbone does not recognize (no backbone match) or will lump under another name (synonyms). See \@ref(verification).

## Read taxa

Read taxa from [`data/interim/taxa_with_verification.csv`](https://github.com/trias-project/unified-checklist/blob/master/data/interim/taxa_with_verification.csv).

```{r verify_taxa-1}
taxa <- read_csv(here("data", "interim", "taxa_with_verification.csv"))
```

## Read verification information {#verification}

Verification information consists of a manually assigned `verificationKey` for taxa the backbone does not recognize (no backbone match) or will lump under another name (synonyms). Those `checklist/backbone/accepted` taxon combinations are stored in [`references/verification.tsv`](https://github.com/trias-project/unified-checklist/blob/master/references/verification.tsv), which was created/updated in previous runs of this chapter and then manually annotated with a `verificationKey`.

```{r verify_taxa-2}
verification <- read_tsv(here("references", "verification.tsv"))
```

## Run and save verification

1. Use the existing verification information to verify taxa with the function `trias::verify_taxa`:

```{r verify_taxa-run_verification, echo = TRUE}
verification <- trias::verify_taxa(taxa, verification)
```

  **Note**: to start a verification from scratch, use `trias::verify_taxa(taxa)`.

2. Save taxa with populated `verificationKey` back to [`data/interim/taxa_with_verification.csv`](https://github.com/trias-project/unified-checklist/blob/master/data/interim/taxa_with_verification.csv).

```{r verify_taxa-3}
verification$taxa %>%
  # Select columns in same order as originally provided
  select(
    taxonKey,
    scientificName,
    taxonID,
    datasetKey,
    nameType,
    issues,
    validDistribution,
    bb_key,
    bb_scientificName,
    bb_kingdom,
    bb_rank,
    bb_taxonomicStatus,
    bb_acceptedKey,
    bb_acceptedName,
    verificationKey
  ) %>%
  write_csv(here("data", "interim", "taxa_with_verification.csv"), na = "")
```

3. Save updated verification back to [`references/verification.tsv`](https://github.com/trias-project/unified-checklist/blob/master/references/verification.tsv) (sorted on `scientificName`)

```{r verify_taxa-4}
verification$verification %>%
  arrange(scientificName) %>%
  write_tsv(here("references", "verification.tsv"), na = "")
```

## Show verification status

### Unmatched taxa

Status | Number of taxa
--- | ---
Verified | `r verification$taxa %>% filter(is.na(bb_key) & !is.na(verificationKey)) %>% nrow()`
Unverified | `r verification$taxa %>% filter(is.na(bb_key) & is.na(verificationKey)) %>% nrow()`
New in this run | `r verification$info$new_unmatched_taxa %>% nrow()`
No longer used | `r verification$info$outdated_unmatched_taxa %>% nrow()`

Unmatched taxa are those for which the automatic backbone matching on the scientific name failed (have [`BACKBONE_MATCH_NONE`](http://gbif.github.io/parsers/apidocs/org/gbif/api/vocabulary/NameUsageIssue.html) in `issues`), either because the scientific name could not be parsed or because it is not listed in the backbone.

Verification (see \@ref(verification)) involves setting the `verificationKey` to a manually found `bb_key`. One could even set multiple `bb_key`s, e.g. to split a hybrid formula into its hybrid parents.

### Synonyms

Status | Number of taxa
--- | ---
Previously verified | `r verification$taxa %>% filter(!is.na(bb_acceptedKey) & !is.na(verificationKey)) %>% nrow()`
Unverified | `r verification$taxa %>% filter(!is.na(bb_acceptedKey) & is.na(verificationKey)) %>% nrow()`
New in this run | `r verification$info$new_synonyms %>% nrow()`
No longer used | `r verification$info$outdated_synonyms %>% nrow()`

Synonyms are taxa that the backbone will lump under an accepted taxon (`taxonomicStatus` = `SYNONYM` or similar). This lumping is often correct, but should be verified.

Verification (see \@ref(verification)) involves setting the `verifiedKey` to one of the following:

* **accepted `bb_key` suggested by GBIF**: backbone synonymy is accepted and taxon will be lumped.
* **another accepted `bb_key`**: backbone synonymy is rejected, but taxon will be lumped under another name.
* **`bb_key` of taxon itself**: backbone synonymy is rejected, taxon will be considered as separate taxon.

### Accepted taxa

Status | Number of taxa
--- | ---
Verified (automatic) | `r verification$taxa %>% filter(bb_taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL") & !is.na(verificationKey)) %>% nrow()`
Unverified |`r verification$taxa %>% filter(bb_taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL") & is.na(verificationKey)) %>% nrow()`

Accepted taxa are those the backbone considers own entities (`taxonomicStatus` = `ACCEPTED` or `DOUBTFUL`) that won't be lumped.

No verification is required for these: `verificationKey` is automatically set to their `bb_key`.

### Erroneous verificationKeys

Provided verification keys that are not found in the GBIF backbone:

```{r verify_taxa-5}
verification$info$check_verificationKey %>%
  filter(!is_from_gbif_backbone) %>%
  distinct(key) %>%
  select(key)
```

### Additional information

Additional information regarding the verification information that is of no further consequence.

* Duplicate taxa (having same `bb_key` and `bb_acceptedKey`) that are quick wins to verify:

```{r verify_taxa-6}
verification$info$duplicates %>%
  group_by(n) %>%
  summarize(
    taxa = n()
  )
```

* Previously listed synonyms for which the scientific name was updated in the backbone:

```{r verify_taxa-7}
verification$info$updated_bb_scientificName
```

* Previously listed synonyms for which the _accepted_ scientific name was updated in the backbone:

```{r verify_taxa-8}
verification$info$updated_bb_acceptedName
```
