# Unify related information

In this chapter we unify related information for each verified taxon.

## Read data

```{r unify_information-1, echo = TRUE}
checklists <- read_csv(here("data", "raw", "checklists.csv"))
taxa <- read_csv(here("data", "interim", "taxa_with_verification.csv"))
distributions <- read_csv(here("data", "raw", "distributions.csv"),
                          col_types = cols(
                            .default = col_character(),
                            taxonKey = col_double()))
speciesprofiles <- read_csv(here("data", "raw", "speciesprofiles.csv"))
descriptions <- read_csv(here("data", "raw", "descriptions.csv"),
                         col_types = cols(
                           .default = col_character(),
                           key = col_double(),
                           taxonKey = col_double()))

# read locality mapping
locations <- read_tsv(here("references", "locality_mapping.tsv"),
                      na = "")
```

## Assign checklist order

1. Get checklist keys as (ordered) vector.

```{r unify_information-2}
checklist_keys <-
  checklists %>%
  pull(datasetKey)
```

2. Assign the checklist order (= index of checklist keys) as an extra column to taxa.

```{r unify_information-3}
taxa <-
  taxa %>%
  rowwise() %>%
  mutate(checklistOrder = which(checklist_keys == datasetKey)[1])
```

## Unify distribution {#unify-distribution}

1. Parse `temporal` (eventDate) information.

```{r unify_information-5, echo = TRUE}
distributions_unified <-
  distributions %>%
  
  # Split temporal on "/" into startYear and endYear
  # If temporal only contains a single year, then endYear will be empty
  separate(
    temporal,
    into = c("startYear", "endYear"),
    sep = "/",
    remove = FALSE,
    convert = TRUE,
    extra = "drop",
    fill = "right"
  ) %>%
  
  # Only keep the first 4 characters: 1968-11-21 -> 1968
  mutate(
    startYear = as.integer(str_sub(startYear, 1, 4)), 
    endYear = as.integer(str_sub(endYear, 1, 4))
  ) %>%
  
  # If endYear is empty (no range), populate it with startYear: 2018 & 2018
  mutate(endYear = if_else(is.na(endYear), startYear, endYear))
```

2. Filter distributions: this was already done in \@ref(filter-on-distribution), where only distributions of interest were kept.

3. Choose a single distribution within a checklist.

**Note**: In contrast to the other extensions, the distribution information can come from multiple taxa within the same checklist. For example:

scientific name | eventDate
--- | ---
[https://www.gbif.org/species/141266662/verbatim](Medicago monantha subsp. incisa (Benth.) Verloove & Lambinon) | 1885/1901
[Medicago monantha subsp. noeana (Boiss.) Greuter & Burdet](https://www.gbif.org/species/141266791/verbatim) | 1886/1955

Both are considered (verified) synonyms of [Medicago monantha Trautv.](https://www.gbif.org/species/2965226), so their distribution information is merged to `1885/1955`. We note both in `taxonKeys` and `scientificNames`.

```{r unify_information-6, echo = TRUE}
distributions_unified <-
  distributions_unified %>%

  # Join distribution with taxon to get verificationKey and checklistOrder
  left_join(taxa, by = "taxonKey") %>%
  
  # Remove records that have no verificationKey (e.g. one wasn't assigned yet)
  filter(!is.na(verificationKey)) %>%
  
  # Group by verificationKey within checklist
  group_by(
    datasetKey,
    checklistOrder,
    verificationKey
  ) %>%
  
  # Take earliest year, latest year and note taxonKey(s) and scientificName(s):
  summarize(
    startYear = as.integer(min(startYear, na.rm = TRUE)),
    endYear = as.integer(max(endYear, na.rm = TRUE)),
    taxonKeys = paste(unique(taxonKey), collapse = "|"),
    scientificNames = paste(unique(scientificName), collapse = "|")
  )
```

4. Choose a single distribution across checklists.

```{r unify_information-7, echo = TRUE}
distributions_unified <-
  distributions_unified %>%
  
  # Sort by checklist order (trustworthiness)
  arrange(checklistOrder) %>%
  
  # Group by verificationKey across checklists
  group_by(verificationKey) %>%
  
  # Select year of most trustworthy checklist (first one)
  # and note that checklist and its taxonKey(s) and scientificName(s)
  summarize(
    startYear = first(startYear),
    endYear = first(endYear),
    datasetKey = first(datasetKey),
    taxonKeys = first(taxonKeys),
    scientificNames = first(scientificNames)
  ) %>%

  # Sort by verificationKey
  arrange(verificationKey)
```

5. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/interim/distributions_unified.csv).

```{r unify_information-8}
write_csv(distributions_unified, here("data", "interim", "distributions_unified.csv"), na = "")
```

## Unify species profiles

1. Filter species profiles.

```{r unify_information-9, echo = TRUE}
speciesprofiles_unified <-
  speciesprofiles %>%
  
  # Remove species profiles that contain NA for any of the attributes
  # This is rare: normally all attributes are populated or there just isn't a
  # species profile for that species
  filter(
    !is.na(marine) &
    !is.na(freshwater) &
    !is.na(terrestrial)
  )
```

2. Choose a single species profile within a checklist.

```{r unify_information-10, echo = TRUE}
speciesprofiles_unified <-
  speciesprofiles_unified %>%

  # Join species profile with taxon to get verificationKey and checklist order
  left_join(taxa, on = taxonKey) %>%
  
  # Remove records that have no verificationKey (e.g. one wasn't assigned yet)
  filter(!is.na(verificationKey)) %>%
  
  # Group by verificationKey within checklist
  group_by(
    datasetKey,
    checklistOrder,
    verificationKey
  ) %>%
  
  # Take first species profile and note taxonKey
  summarize(
    marine = first(marine),
    freshwater = first(freshwater),
    terrestrial = first(terrestrial),
    taxonKey = first(taxonKey),
    scientificName = first(scientificName)
  )
```

3. Choose a single species profile across checklists.

```{r unify_information-11, echo = TRUE}
speciesprofiles_unified <-
  speciesprofiles_unified %>%
  
  # Sort by checklist order (trustworthiness)
  arrange(checklistOrder) %>%
  
  # Group by verificationKey across checklists
  group_by(verificationKey) %>%
  
  # Select species profile of most trustworthy checklist (first one)
  # and note that checklist and its taxonKey
  summarize(
    marine = first(marine),
    freshwater = first(freshwater),
    terrestrial = first(terrestrial),
    datasetKey = first(datasetKey),
    taxonKey = first(taxonKey),
    scientificName = first(scientificName)
  ) %>%

  # Sort by verificationKey
  arrange(verificationKey)
```

4. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/interim/speciesprofiles_unified.csv).

```{r unify_information-12}
write_csv(speciesprofiles_unified, here("data", "interim", "speciesprofiles_unified.csv"), na = "")
```

## Unify descriptions

1. Filter non `NA` descriptions.

```{r unify_information-13}
descriptions_unified <-
  descriptions %>%
  
  # Filter on non-NA descriptions
  filter(
    !is.na(description)
  )
```

2. Select unique descriptions (within their type) within a checklist.

```{r unify_information-14, echo = TRUE}
descriptions_unified <-
  descriptions_unified %>%

  # Join species profile with taxon to get verificationKey and checklist order
  left_join(taxa, on = taxonKey) %>%
  
  # Remove records that have no verificationKey (e.g. one wasn't assigned yet)
  filter(!is.na(verificationKey)) %>%
  
  # Group by type and verificationKey within checklist
  group_by(
    datasetKey,
    checklistOrder,
    verificationKey,
    type,
    description
  ) %>%
  
  # Take first taxonKey and scientificName
  summarize(
    taxonKey = first(taxonKey),
    scientificName = first(scientificName)
  )
```

3. Select unique descriptions (within their type) across checklists.

```{r unify_information-15, echo = TRUE}
descriptions_unified <-
  descriptions_unified %>%
  
  # Sort by checklist order (trustworthiness)
  arrange(checklistOrder) %>%
  
  # Group by type and verificationKey across checklists
  group_by(
    type,
    description,
    verificationKey
  ) %>%
  
  # Select first datasetKey, taxonKey and scientificName
  summarize(
    datasetKey = first(datasetKey),
    taxonKey = first(taxonKey),
    scientificName = first(scientificName)
  ) %>%

  # Move verificationKey to beginning and drop checklist_order
  select(verificationKey, everything()) %>%
  
  # Sort by verificationKey and type
  arrange(verificationKey, type)
```

4. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/interim/descriptions_unified.csv).

```{r unify_information-16}
write_csv(descriptions_unified, here("data", "interim", "descriptions_unified.csv"), na = "")
```
