# Darwin Core mapping

In this chapter we standardize the unified information to a Darwin Core checklist that can be harvested by GBIF.

## Read unified data

```{r echo = TRUE}
checklists <- read_csv(here("data", "raw", "checklists.csv"), col_types = "ccTcccc")
input_taxa <- read_csv(here("data", "interim", "taxa_unified.csv"), col_types = "icicccccccccicccccciccc")
input_distributions <- read_csv(here("data", "interim", "distributions_unified.csv"), col_types = "iiicic")
input_speciesprofiles <- read_csv(here("data", "interim", "speciesprofiles_unified.csv"), col_types = "illlcic")
input_descriptions <- read_csv(here("data", "interim", "descriptions_unified.csv"), col_types = "icccic")
```

## Preview data

1. Number of rows per file and corresponding mapping section in this chapter:

File | Number of rows
--- | ---
taxa | `r nrow(input_taxa)`
distributions | `r nrow(input_distributions)`
speciesprofiles | `r nrow(input_speciesprofiles)`
descriptions | `r nrow(input_descriptions)`

2. Number of taxa per kingdom and rank:

```{r}
input_taxa %>%
  group_by(kingdom, rank) %>%
  summarize(
    `taxa` = n()
  ) %>%
  adorn_totals("row")
```

3. Number of taxa and descriptions per type:

```{r}
input_descriptions %>%
  group_by(type) %>%
  summarize(
    `taxa` = n(),
    `unique taxa` = n_distinct(verificationKey),
    `unique descriptions` = n_distinct(description)
  ) %>%
  adorn_totals("row")
```

## How we cite our sources {#citing-sources}

Each row of information in the Taxon core and the extensions is based on a specific source:

File | Source | Field for citation | Mapping section
--- | --- | --- | ---
Taxon core | a taxon in the [GBIF Backbone Taxonomy](https://doi.org/10.15468/39omei) | `bibliographicCitation`| \@ref(taxon-core)
Distribution extension | a taxon in a source checklist | `source` | \@ref(distribution-extension)
Species profile extension | a taxon in a source checklist | `source` | \@ref(species-profile-extension)
Description extension | a taxon in a source checklist | `source` | \@ref(description-extension)

To reference this source, we will use the **GBIF citation format for species pages**, prefixed with the URL of that page. E.g. for the distribution of _Nymphea marliacea Marliac_ this would be:

> https://www.gbif.org/species/141264581: Nymphaea marliacea Marliac in Verloove F, Groom Q, Brosens D, Desmet P, Reyserhove L (2018). Manual of the Alien Plants of Belgium. Version 1.7. Botanic Garden Meise. Checklist dataset https://doi.org/10.15468/wtda1m.

This information is a combination of: 

- `taxonKey`: e.g. `1412645812` (contained in `distributions.csv`),
- `scientificName`: e.g. `Nymphaea marliacea Marliac` (contained in `distributions.csv`),
- `citation`: e.g. `Verloove F, Groom Q, Brosens D, Desmet P, Reyserhove L (2018). Manual of the Alien Plants of Belgium. Version 1.7. Botanic Garden Meise. Checklist dataset https://doi.org/10.15468/wtda1m.` (contained in `checklists.csv`)

To generate this full citation, we create a helper function `add_source_citation(df, new_column, dataset_info)`.

```{r}
add_source_citation <- function(df, new_column, dataset_info,
                                dataset_key = "datasetKey",
                                taxon_key = "taxonKey",
                                scientific_name = "scientificName") {
  df %>%
  
  # Join df with dataset_info
  left_join(
    select(dataset_info, datasetKey, citation),
    by = setNames("datasetKey", dataset_key)
  ) %>%

  # Build full citation in new_column (requires !! to bypass nse)
  mutate(!!new_column := paste0(
    "https://www.gbif.org/species/",
    df[[taxon_key]],
    ": ",
    df[[scientific_name]],
    " in ",
    citation
  )) %>%
  
  # Remove added citation field
  select(-citation)
}
```

## Taxon core {#taxon-core}

### Pre-processing

1. Create a dataframe `taxon` from the unified taxa.

```{r}
taxon <- input_taxa
```

2. Separate `canonicalName` in `canonicalName_genus`, `canonicalName_species` and `canonicalName_infraspecific` (on whitespace).

```{r}
taxon %<>% separate(
  canonicalName,
  into = c(
    "canonicalName_genus",
    "canonicalName_species",
    "canonicalName_infraspecific"
  ),
  sep = "\\s+", # Whitespace
  remove = FALSE,
  convert = TRUE,
  extra = "drop",
  fill = "right"
)
```

### Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

#### language

```{r echo = TRUE}
taxon %<>% mutate(dwc_language = "en")
```

#### license

The license under which (each record of) the unified checklist will be published should be the most restrictive license of the source checklists. The **potential licenses** are the three Creative Commons licenses [supported by GBIF](https://www.gbif.org/terms) (ordered from least to most restrictive):

```{r}
legal_licenses <- tibble::tibble(
  license = c(
    "http://creativecommons.org/publicdomain/zero/1.0/legalcode",
    "http://creativecommons.org/licenses/by/4.0/legalcode",
    "http://creativecommons.org/licenses/by-nc/4.0/legalcode"
    ),
  ranking = c(1:3)
)
legal_licenses
```

The **actual licenses** of the source checklists _and_ the [GBIF Backbone Taxonomy](https://doi.org/10.15468/39omei) (which we use for the taxon core) are:

```{r}
checklist_licenses <-
  checklists %>%
  group_by(license) %>%
  count() %>%
  ungroup()
checklist_licenses
```

Based on the above `ranking`, the most restrictive license is:

```{r}
most_restrictive_license <-
  checklist_licenses %>%
  left_join(legal_licenses, by = "license") %>%
  filter(ranking == max(ranking)) %>%
  pull(license)
most_restrictive_license
```

Which we use for our `license`:

```{r echo = TRUE}
taxon %<>% mutate(dwc_license = most_restrictive_license)
```

#### rightsHolder

We do not set `rightsHolder` as the taxon and its related information is based on different source checklists (which in turn are based on other sources), published by different organizations, and mostly released under CC0. Instead, we make an effort to cite the sources (see \@ref(citing-sources)).

```{r echo = TRUE}
taxon %<>% mutate(dwc_rightsHolder = NA)
```

#### bibliographicCitation

See \@ref(citing-sources):

```{r echo = TRUE}
# Add temporary field with datasetKey of GBIF Backbone Taxonomy
# taxon %<>% mutate(datasetKey = "d7dddbf4-2cf0-4f39-9b2a-bb099caae36c")

taxon %<>% add_source_citation(
  new_column = "dwc_bibliographicCitation",
  dataset_info = checklists,
  taxon_key = "verificationKey"
)
```

#### datasetID

```{r echo = TRUE}
taxon %<>% mutate(dwc_datasetID = "")
```

#### institutionCode

```{r echo = TRUE}
taxon %<>% mutate(dwc_institutionCode = "ISSG") # Invasive Species Specialist Group ISSG
```

#### datasetName

```{r echo = TRUE}
taxon %<>% mutate(dwc_datasetName = "Global Register of Introduced and Invasive Species - Belgium")
```

#### references

URL of the GBIF Backbone Taxonomy taxon on gbif.org:

```{r echo = TRUE}
taxon %<>% mutate(dwc_references = paste0("https://www.gbif.org/species/", verificationKey))
```

#### taxonID

URL of the GBIF Backbone Taxonomy taxon on gbif.org:

```{r echo = TRUE}
taxon %<>% mutate(dwc_taxonID = paste0("https://www.gbif.org/species/", verificationKey))
```

#### acceptedNameUsageID

```{r echo = TRUE}
taxon %<>% mutate(dwc_acceptedNameUsageID = case_when(
  is.na(acceptedKey) ~ paste0("https://www.gbif.org/species/", verificationKey), # If empty acceptedKey, use verificationKey
  TRUE ~ paste0("https://www.gbif.org/species/", acceptedKey)
))
```

#### scientificName

```{r echo = TRUE}
taxon %<>% mutate(dwc_scientificName = scientificName)
```

#### acceptedNameUsage

```{r echo = TRUE}
taxon %<>% mutate(dwc_acceptedNameUsage = case_when(
  is.na(acceptedName) ~ scientificName, # If empty acceptedName, use scientificName
  TRUE ~ acceptedName
))
```

#### kingdom

```{r echo = TRUE}
taxon %<>% mutate(dwc_kingdom = kingdom)
```

#### phylum

```{r echo = TRUE}
taxon %<>% mutate(dwc_phylum = phylum)
```

#### class

```{r echo = TRUE}
taxon %<>% mutate(dwc_phylum = class)
```

#### order

```{r echo = TRUE}
taxon %<>% mutate(dwc_order = order)
```

#### family

```{r echo = TRUE}
taxon %<>% mutate(dwc_family = family)
```

#### genus

`genus` is part of the higher classification, which is provided by the GBIF Backbone Taxonomy. We will use that, but note that for some synonyms it might differ from the genus in the scientific name:

```{r}
taxon %>%
  filter(genus != canonicalName_genus) %>%
  select(verificationKey, scientificName, genus, canonicalName_genus, taxonomicStatus)
```

```{r echo = TRUE}
taxon %<>% mutate(dwc_genus = genus)
```

#### specificEpithet

```{r echo = TRUE}
taxon %<>% mutate(dwc_specificEpithet = canonicalName_species)
```

#### infraspecificEpithet

```{r echo = TRUE}
taxon %<>% mutate(dwc_infraspecificEpithet = canonicalName_infraspecific)
```

#### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(rank) %>%
  count()
```

Map values as is, in UPPERCASE, so it is clearer this information comes from the GBIF Backbone Taxonomy:

```{r echo = TRUE}
taxon %<>% mutate(dwc_taxonRank = rank)
```

#### scientificNameAuthorship

```{r echo = TRUE}
taxon %<>% mutate(dwc_scientificNameAuthorship = authorship)
```

#### taxonomicStatus

Inspect values:

```{r}
taxon %>%
  group_by(taxonomicStatus) %>%
  count()
```

Map values as is, in UPPERCASE, so it is clearer this information comes from the GBIF Backbone Taxonomy:

```{r echo = TRUE}
taxon %<>% mutate(dwc_taxonomicStatus = taxonomicStatus)
```

#### taxonRemarks

Here we list the checklists that were _considered_ for unifying information about a taxon, i.e. the checklists we selected (see \@ref(choose-checklists)) in which the taxon appeared and got through verification (see \@ref(verification)). In the case of multiple considered checklists, it is possible that not all of them are selected as a source (see \@ref(citing-sources)) when unifying information.

The `datasetKey`s of the considered checklists are listed in `datasetKeys`. Since we want the DOI instead, we will separate this information into columns, gather to rows and join with the DOI, and spread and combine again into a single column.

```{r}
taxon <-
  taxon %>%
  
  # Separate _datasetKeys on "," in as many columns as there are checklists
  separate(
    datasetKeys,
    into = c(paste("source", 1:nrow(checklists), sep = "_")),
    sep = ",",
    remove = FALSE,
    fill = "right"
  ) %>%

  # Gather to one row per source (multiple rows)
  gather(
    key = source,
    value = key,
    paste("source", 1:nrow(checklists), sep = "_"),
    na.rm = TRUE,
    convert = FALSE
  ) %>%

  # Join with checklists
  left_join(
    select(checklists, datasetKey, doi),
    by = c("key" = "datasetKey")
  ) %>%
  
  # Remove column "key"
  select(-key) %>%
  
  # Spread back to one row per taxon (multiple columns)
  spread(source, doi) %>%
  
  # Combine source columns into one column "datasetDOIs"
  unite(
    col = "datasetDOIs",
    starts_with("source_"),
    sep = ", ",
    remove = TRUE
  ) %>%
  
  # Delete "NA" string values
  mutate(datasetDOIs = str_remove_all(datasetDOIs, "(, NA)"))
```
  
Map to `taxonRemarks`:

```{r echo = TRUE}
taxon %<>% mutate(dwc_taxonRemarks = paste(
  "Sources considered for unifying information: ", datasetDOIs
))
```

### Post-processing

1. Only keep the Darwin Core columns.
  
```{r}
taxon %<>% select(starts_with("dwc_"))
```
  
2. Drop the `dwc_` prefix.
  
```{r}
colnames(taxon) <- str_replace(colnames(taxon), "dwc_", "")
```

3. Sort on `taxonID`.

```{r}
taxon %<>% arrange(taxonID)
```

4. Preview data:

```{r}
taxon %>% head()
```

5. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/processed/taxon.csv).

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```

## Distribution extension {#distribution-extension}

### Pre-processing

Create a dataframe `distribution` from the unified distributions.

```{r}
distribution <- input_distributions
```

### Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml). Because of the scope (see \@ref(filter-on-distribution)) of the dataset, we can set all distributions to `occurrenceStatus:present` and `establishmentMeans:introduced` in `Belgium`.

#### taxonID

```{r echo = TRUE}
distribution %<>% mutate(dwc_taxonID = paste0("https://www.gbif.org/species/", verificationKey))
```

#### locationID

```{r echo = TRUE}
distribution %<>% mutate(dwc_locationID = "ISO_3166-2:BE") 
```

#### locality

```{r echo = TRUE}
distribution %<>% mutate(dwc_locality = "Belgium") 
```

#### countryCode

```{r echo = TRUE}
distribution %<>% mutate(dwc_countryCode = "BE") 
```

#### occurrenceStatus

```{r echo = TRUE}
distribution %<>% mutate(dwc_occurrenceStatus = "present") 
```

#### establishmentMeans

```{r echo = TRUE}
distribution %<>% mutate(dwc_establishmentMeans = "introduced") 
```

#### eventDate

The distribution information applies to a certain date range, which we will express here as an ISO 8601 date `yyyy/yyyy` (`startYear/endYear`). How the information for `startYear` and `endYear` is extracted from the source checklists, is described in \@ref(unify-distribution). As a result, each taxon in `input_distribution` has _or_ an `startYear` and an `endYear` _or_ no `eventDate` information at all.

```{r echo = TRUE}
distribution %<>% mutate(dwc_eventDate = case_when(
  is.na(startYear) & is.na(endYear) ~ "",
  TRUE ~ paste(startYear, endYear, sep = "/")
))
``` 
 
#### source

See \@ref(citing-sources):

```{r echo = TRUE}
distribution %<>% add_source_citation(
  new_column = "dwc_source",
  dataset_info = checklists
)
```

### Post-processing

1. Only keep the Darwin Core columns.
  
```{r}
distribution %<>% select(starts_with("dwc_"))
```
  
2. Drop the `dwc_` prefix.
  
```{r}
colnames(distribution) <- str_replace(colnames(distribution), "dwc_", "")
```

3. Sort on `taxonID`.

```{r}
distribution %<>% arrange(taxonID)
```

4. Preview data:

```{r}
distribution %>% head()
```

5. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/processed/distribution.csv).

```{r}
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
```

## Species profile extension {#species-profile-extension}

Create a dataframe `species_profile` from the unified species profiles.

```{r}
species_profile <- input_speciesprofiles
```

### Term mapping

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

#### taxonID

```{r echo = TRUE}
species_profile %<>% mutate(dwc_taxonID = paste0("https://www.gbif.org/species/", verificationKey))
```

#### isMarine

```{r echo = TRUE}
species_profile %<>% mutate(dwc_isMarine = marine)
```

#### isFreshwater

```{r echo = TRUE}
species_profile %<>% mutate(dwc_isFreshwater = freshwater)
```

#### isTerrestrial

```{r echo = TRUE}
species_profile %<>% mutate(dwc_isTerrestrial = terrestrial)
```

#### isInvasive

The source checklists currently do not include information on the invasive nature of the taxa. We plan to add that information in an update of the dataset.

```{r echo = TRUE}
species_profile %<>% mutate(dwc_isInvasive = "")
```

#### habitat

ISSG also used the field `habitat`, in which we will summarize the information from `isMarine`, `isFreshwater` and `isTerrestrial`. 

```{r}
species_profile %>%
  group_by(marine, freshwater, terrestrial) %>% 
  summarize(records = n()) %>% 
  arrange(marine, freshwater, terrestrial)
```

Map `habitat`:

```{r echo = TRUE}
species_profile %<>% mutate(dwc_habitat = case_when(
  marine == "FALSE" & freshwater == "FALSE" & terrestrial == "TRUE"  ~ "terrestrial",
  marine == "FALSE" & freshwater == "TRUE"  & terrestrial == "FALSE" ~ "freshwater",
  marine == "FALSE" & freshwater == "TRUE"  & terrestrial == "TRUE"  ~ "freshwater|terrestrial",
  marine == "TRUE"  & freshwater == "FALSE" & terrestrial == "FALSE" ~ "marine",
  marine == "TRUE"  & freshwater == "FALSE" & terrestrial == "TRUE"  ~ "marine|terrestrial",
  marine == "TRUE"  & freshwater == "TRUE"  & terrestrial == "FALSE" ~ "marine|freshwater",
  marine == "TRUE"  & freshwater == "TRUE"  & terrestrial == "TRUE"  ~ "marine|freshwater|terrestrial"
))
```

Show mapping:

```{r}
species_profile %>% 
  group_by(dwc_isMarine, dwc_isFreshwater, dwc_isTerrestrial, dwc_habitat) %>% 
  summarize(records = n()) 
```

#### source

See \@ref(citing-sources):

```{r echo = TRUE}
species_profile %<>% add_source_citation(
  new_column = "dwc_source",
  dataset_info = checklists
)
```

### Post-processing

1. Only keep the Darwin Core columns.
  
```{r}
species_profile %<>% select(starts_with("dwc_"))
```
  
2. Drop the `dwc_` prefix.
  
```{r}
colnames(species_profile) <- str_replace(colnames(species_profile), "dwc_", "")
```

3. Sort on `taxonID`.

```{r}
species_profile %<>% arrange(taxonID)
```

4. Preview data:

```{r}
species_profile %>% head()
```

5. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/processed/speciesprofile.csv).

```{r}
write.csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
```

## Description extension {#description-extension}

### Pre-processing

Create a dataframe `description` from the unified descriptions.

```{r}
description <- input_descriptions
```

### Term mapping

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

#### taxonID

```{r echo = TRUE}
description %<>% mutate(dwc_taxonID = paste0("https://www.gbif.org/species/", verificationKey)) 
```

#### description

```{r echo = TRUE}
description %<>% mutate(dwc_description = description)
```

#### type

```{r echo = TRUE}
description %<>% mutate(dwc_type = type)
```

#### language

```{r echo = TRUE}
description %<>% mutate(dwc_language = "en")
```

#### source

See \@ref(citing-sources):

```{r echo = TRUE}
description %<>% add_source_citation(
  new_column = "dwc_source",
  dataset_info = checklists
)
```

### Post-processing

1. Only keep the Darwin Core columns.
  
```{r}
description %<>% select(starts_with("dwc_"))
```
  
2. Drop the `dwc_` prefix.
  
```{r}
colnames(description) <- str_replace(colnames(description), "dwc_", "")
```

3. Sort on `taxonID`.

```{r}
description %<>% arrange(taxonID)
```

4. Preview data:

```{r}
description %>% head()
```

5. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/processed/description.csv).

```{r}
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
