# GRIIS mapping

In this chapter we export the unified information to an Excel template that can be used by the Global Register of Introduced and Invasive Species (GRIIS).

## Read Darwin Core data

```{r echo = TRUE}
dwc_taxon <- read_csv(here("data", "processed", "taxon.csv"))
dwc_distribution <- read_csv(here("data", "processed", "distribution.csv"))
dwc_speciesprofile <- read_csv(here("data", "processed", "speciesprofile.csv"))
```

Select only national distributions

```{r}
dwc_distribution <- dwc_distribution %>% filter(locality == "Belgium")
```

## GRIIS format

### Pre-processing

1. Check if the taxon core, distribution extension and species profile extension contain only one row per `taxonID`.

```{r}
nrow(dwc_taxon) == dwc_taxon %>% distinct(taxonID) %>% nrow()
nrow(dwc_distribution) == dwc_distribution %>% distinct(taxonID) %>% nrow()
nrow(dwc_speciesprofile) == dwc_speciesprofile %>% distinct(taxonID) %>% nrow()
```

2. Join the 3 files into one dataframe `griis`.

```{r}
griis <-
  # taxa
  dwc_taxon %>%
  
  # join with distributions
  left_join(
    dwc_distribution,
    on = "taxonID"
  ) %>%
  
  # join species profiles
  left_join(
    dwc_speciesprofile,
    on = "taxonID"
  )
```

### Term mapping

#### taxonID

```{r echo = TRUE}
griis %<>% mutate(griis_taxonID = taxonID)
```

#### countryCode

```{r echo = TRUE}
griis %<>% mutate(griis_countryCode = countryCode)
```

#### island

```{r echo = TRUE}
griis %<>% mutate(griis_island = "")
```

#### scientificName

```{r echo = TRUE}
griis %<>% mutate(griis_scientificName = scientificName)
```

#### acceptedNameUsage

```{r echo = TRUE}
griis %<>% mutate(griis_acceptedNameUsage = case_when(
  acceptedNameUsage != scientificName ~ acceptedNameUsage,
  TRUE ~ ""
))
```

#### taxonRank

```{r echo = TRUE}
griis %<>% mutate(griis_taxonRank = taxonRank)
```

#### taxonomicStatus

```{r echo = TRUE}
griis %<>% mutate(griis_taxonomicStatus = taxonomicStatus)
```

#### kingdom

```{r echo = TRUE}
griis %<>% mutate(griis_kingdom = kingdom)
```

#### phylum

```{r echo = TRUE}
griis %<>% mutate(griis_phylum = phylum)
```

#### class

```{r echo = TRUE}
griis %<>% mutate(griis_class = class)
```

#### order

```{r echo = TRUE}
griis %<>% mutate(griis_order = order)
```

#### family

```{r echo = TRUE}
griis %<>% mutate(griis_family = family)
```

#### habitat

```{r echo = TRUE}
griis %<>% mutate(griis_habitat = habitat)
```

#### occurrenceStatus

```{r echo = TRUE}
griis %<>% mutate(griis_occurrenceStatus = occurrenceStatus)
```

#### establishmentMeans

```{r echo = TRUE}
griis %<>% mutate(griis_establishmentMeans = recode(establishmentMeans,
  "introduced" = "alien",
  .default = "",
  .missing = ""
))
```

#### isInvasive

```{r echo = TRUE}
griis %<>% mutate(griis_isInvasive = case_when(
  is.na(isInvasive) ~ "Null"
))
```

### eventDate

```{r echo = TRUE}
griis %<>% mutate(griis_eventDate = eventDate)
```

## Post-processing

1. Only keep the GRIIS columns.
  
```{r}
griis %<>% select(starts_with("griis_"))
```
  
2. Drop the `griis_` prefix.
  
```{r}
colnames(griis) <- str_remove(colnames(griis), "griis_")
```

3. Sort on `scientificName` (the default for GRIIS lists).

```{r}
griis %<>% arrange(scientificName)
```

4. Preview data:

```{r}
griis %>% head()
```

5. Save to [Excel](https://github.com/trias-project/unified-checklist/blob/master/data/processed/Belgium_GBIF_GRIIS.xlsx).

```{r}
openxlsx::write.xlsx(griis, here("data", "processed", "Belgium_GBIF_GRIIS.xlsx"))
```
