# Get synonyms of all accepted taxa

In this chapter we retrieve all synonyms of the taxa in the unified checklist. Together with the accepted names, such a list allows to search for non-native scientific names in e.g. occurrence databases.

## Get unified taxa

1. Read `data/interim/taxa_unified.csv`.

```{r get_synonyms-1}
input_taxa <- read_csv(here("data", "interim", "taxa_unified.csv"))
```

2. Define columns of interest.

```{r get_synonyms-2}
selected_columns <- c(
  "key",
  "scientificName",
  "canonicalName",
  "authorship",
  "rank",
  "taxonomicStatus",
  "kingdom",
  "phylum",
  "class",
  "order",
  "family",
  "genus",
  "species",
  "parentKey",
  "parent",
  "acceptedKey",
  "accepted"
)
```

3. Select/rename columns of interest.

```{r get_synonyms-3}
input_taxa <-
  input_taxa %>%
  rename(key = verificationKey) %>%
  rename(accepted = acceptedName) %>%
  select(all_of(selected_columns))
```

## Find synonyms

1. Retrieve all synonyms (from the GBIF Backbone Taxonomy) of the taxa in our list. Synonym taxa won't return results, but are fine to leave in the list. _Note: this step can take few minutes._

```{r get_synonyms-get-synonyms-from-gbif}
progress_bar <- progress_estimated(nrow(input_taxa))

detach("package:tidylog") # to remove info interrupting progress bar
synonyms <- map_dfr(
  input_taxa$key, 
  function(key) {
    progress_bar$tick()$print()
      name_usage(
        key = key, 
        data = "synonyms")$data %>%
      select(-contains("issues"))
  }
)
library(tidylog)
```

2. Select columns of interest in synonyms.

```{r get_synonyms-4}
synonyms <-
  synonyms %>%
  select(all_of(selected_columns))
```

3. Merge input taxa with their synonyms.

```{r get_synonyms-5}
taxa_and_synonyms <-
  input_taxa %>%
  mutate(source = "unified checklist") %>%
  union_all(
    synonyms %>% mutate(source = "gbif synonym")
  )
```

4. Save as [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/interim/taxa_unified_and_synonyms.csv).

```{r get_synonyms-6}
write_csv(taxa_and_synonyms, here("data", "interim", "taxa_unified_and_synonyms.csv"), na = "")
```
