# Read source data

Create data frame `input_data` from the raw and interim data:

```{r}
checklists <- read_csv(here("data", "raw", "checklists.csv"))
input_taxa <- read_csv(here("data", "interim", "taxa_unified.csv"))
input_distributions <- read_csv(here("data", "interim", "distributions_unified.csv"))
input_speciesprofiles <- read_csv(here("data", "interim", "speciesprofiles_unified.csv"))
input_descriptions <- read_csv(here("data", "interim", "descriptions_unified.csv"))
```

# Process data

Add prefix `input_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(input_taxa) <- paste0("input_", colnames(input_taxa))
colnames(input_distributions) <- paste0("input_", colnames(input_distributions))
colnames(input_speciesprofiles) <- paste0("input_", colnames(input_speciesprofiles))
colnames(input_descriptions) <- paste0("input_", colnames(input_descriptions))
```

# Preview data

1. Number of rows:

* Taxa: `r nrow(input_taxa)`
* Distributions: `r nrow(input_distributions)`
* Species profiles: `r nrow(input_speciesprofiles)`
* Descriptions: `r nrow(input_descriptions)`

2. Number of taxa per kingdom and rank:

```{r echo = FALSE}
input_taxa %>%
  group_by(input_kingdom, input_rank) %>%
  summarize(
    `taxa` = n()
  ) %>%
  adorn_totals("row")
```

3. Number of taxa and descriptions per type:

```{r echo = FALSE}
input_descriptions %>%
  group_by(input_type) %>%
  summarize(
    `taxa` = n(),
    `unique taxa` = n_distinct(input_verificationKey),
    `unique descriptions` = n_distinct(input_description)
  ) %>%
  adorn_totals("row")
```

# Taxon core

## Pre-processing

1. Create a dataframe from the source taxa:

```{r}
taxon <- input_taxa
```

2. Separate `input_canonicalName` in 3 parts:

```{r}
taxon %<>% separate(
  input_canonicalName,
  into = c(
    "input_canonicalName_genus",
    "input_canonicalName_species",
    "input_canonicalName_infraspecific"
  ),
  sep = "\\s+",
  remove = FALSE,
  convert = TRUE,
  extra = "drop",
  fill = "right"
)
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

The license of the unified checklist equals the least open license of the source checklists. In the TrIAS project, the checklists can be published under the following licenses, ranked from most open (1) to more restricive (3):

```{r}
legal_licenses <- tibble::tibble(
  license = c(
    "http://creativecommons.org/publicdomain/zero/1.0/legalcode",
    "http://creativecommons.org/licenses/by/4.0/legalcode",
    "https://creativecommons.org/licenses/by-nc/4.0/legalcode"
    ),
  ranking = c(1:3)
)
legal_licenses
```

Extract all unique licenses from the source checklists:

```{r}
checklists_licenses <- checklists %>% distinct(license)
checklists_licenses
```

We need to consider the license of the [gbif backbone taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c) as well, which is a CC-BY license:

```{r}
checklists_licenses %<>% add_row(license = "http://creativecommons.org/licenses/by/4.0/legalcode")
```

We add the ranking to the checklist licenses and extract the most restrictive one by its highest ranking:

```{r}
most_restrictive_license <- checklists_licenses %>% 
  left_join(legal_licenses, by = "license") %>% 
  filter(ranking == max(ranking)) %>% 
  select(license)
most_restrictive_license
```

Add this to the taxon core:

```{r}
taxon %<>% mutate(license = sQuote(most_restrictive_license))
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = NA)
```

### bibliographicCitation

```{r}
#include?
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### institutionCode

```{r}
taxon %<>% mutate(institutionCode = "ISSG")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Unified checklist of alien species in Belgium")
```

### references

```{r}
taxon %<>% mutate(references = paste0("https://www.gbif.org/species/", input_verificationKey))
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = paste0("https://www.gbif.org/species/", input_verificationKey))
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = input_scientificName)
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = input_kingdom)
```

### phylum

```{r}
taxon %<>% mutate(phylum = input_phylum)
```

### class

```{r}
taxon %<>% mutate(phylum = input_class)
```

### order

```{r}
taxon %<>% mutate(order = input_order)
```

### family

```{r}
taxon %<>% mutate(family = input_family)
```

### genus

Inspect differences between `input_genus` and `input_canonicalName_genus`:

```{r}
taxon %>%
  filter(input_genus != input_canonicalName_genus) %>%
  select(input_verificationKey, input_scientificName, input_genus, input_canonicalName_genus, input_taxonomicStatus)
```

The `input_genus` under which the GBIF Backbone Taxonomy has classified a taxon can differ from the genus in the scientific name. Since we consider all our taxa as accepted taxa, we take the genus from the scientific name (= `input_canonicalName_genus`).

```{r}
taxon %<>% mutate(genus = input_canonicalName_genus)
```

### specificEpithet

```{r}
taxon %<>% mutate(specificEpithet = input_canonicalName_species)
```

### infraspecificEpithet

```{r}
taxon %<>% mutate(infraspecificEpithet = input_canonicalName_infraspecific)
```

### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(input_rank) %>%
  count()
```

Map values to lowercase:

```{r}
taxon %<>% mutate(taxonRank = str_to_lower(input_rank))
```

### scientificNameAuthorship

```{r}
taxon %<>% mutate(scientificNameAuthorship = input_authorship)
```

### taxonRemarks

Here, we integrate the reference (doi) of the *source* of the taxon information (contained in `input_sourceChecklists`), i.e. the link to the specific taxon in the specific checklist.

`input_sourceChecklist` now integrates all keys of the original checklists (sources) that include the specific taxon, separated by a comma. We need to split these keys and transform `taxon` such that one row contains only one sourceChecklist:

```{r}
# Split `input_sourceChecklist` based on ",". The maximal amount of generated columns equals the total amount of source checklists in `checklists:

taxon %<>% separate(input_sourceChecklists, 
                    into = c(paste("source", 1:nrow(checklists), sep = "_")), 
                    sep =",", 
                    fill = "right") 

# Create `taxon` with one row containing only one sourceChecklist:

taxon %<>% gather(
  key = source, 
  value = key,
  paste("source", 1:nrow(checklists), sep = "_"),
  na.rm = TRUE, 
  convert = FALSE)

# Order by `input_verificationKey`:

taxon %<>% arrange(input_verificationKey)
```
  
Next, We combine the keys of the original sources with their respective doi's using `checklists`:

```{r}
taxon %<>% left_join(
  select(checklists, datasetKey, doi),
  by = c("key" = "datasetKey"))
```

Transform `taxon` to wide dataset, with one row containing the information from one `verificationKey`:

```{r}
# First remove column `key` (information is now in `doi`)
taxon %<>% select(-key)

# Transform taxon to wide dataset, with one column containing one doi of the source checklist:
taxon %<>% spread(source, doi)
```

Generate `taxonRemarks` by combining information in the source columns:

```{r}
# Combine source information in one column:
taxon %<>% unite(col = "source", starts_with("source_"), sep = ",")

# Delete "NA":
taxon %<>% mutate(source = str_remove_all(source,"(,NA)"))

# Add "Sources considered for unifying information: "
taxon %<>% mutate(taxonRemarks = paste("Sources considered for unifying information:", source, sep = " "))
```

## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("input_"), -source)
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```

# Create source citation function

All records in the extensions will be documented with a source, i.e. the link to the specific checklist taxon that was used for the specific piece of information. We build this function here first.

```{r}
build_source_citation <- function(extension, checklists){
  
  # Join the extension and source_citation by their datasetKey:
  
  extension <-
    extension %>% 
    left_join(
      select(checklists, datasetKey, citation),
      by = setNames("datasetKey", "input_datasetKey")
    ) 
  
  # Generate the field `source` by combining `taxonKey`, `scientificName` and `citation`. This is an example of how the final source should look like: 
  # https://www.gbif.org/species/141264581: Nymphaea marliacea Marliac in Verloove F, Groom Q, Brosens D, Desmet P, Reyserhove L (2018). Manual of the Alien Plants of Belgium. Version 1.7. Botanic Garden Meise. Checklist dataset https://doi.org/10.15468/wtda1m
  
    extension %>% 
    mutate(source = paste0(
      "https://www.gbif.org/species/",
      input_taxonKey,
      ": ",
      input_scientificName,
      " in ",
      citation,
      sep = " "
    )) %>% 
    select(-citation)           # Remove citation from the dataframe
    }
```

# Distribution extension

## Pre-processing

Create a dataframe with all data:

```{r}
distribution <- input_distributions
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

## Term mapping

### taxonID

```{r}
distribution %<>% mutate(taxonID = paste0("https://www.gbif.org/species/", input_verificationKey))
```

### locationID

```{r}
distribution %<>% mutate(locationID = "ISO_3166-2:BE") 
```

### locality

```{r}
distribution %<>% mutate(locatlity = "Belgium") 
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = "BE") 
```

### occurrenceStatus

```{r}
distribution %<>% mutate(occurrenceStatus = "present") 
```

### establishmentMeans

```{r}
distribution %<>% mutate(establishmentMeans = "introduced") 
```

### eventDate 

The distribution information applies to a certain date range, which we will express here as an ISO 8601 date `yyyy/yyyy` (`input_startYear/input_endYear`). How the information for `input_startYear` and `input_endYear` is extracted from the TrIAS species checklists, is described [here](https://trias-project.github.io/unified-checklist/unify_information.html#unify-distribution). As a result, each taxon in `input_distribution` has *or* an `input_startYear` and an `input_endYear` *or* no eventDate information at all.

```{r}
distribution  %<>% mutate(eventDate = case_when(
  is.na(input_startYear) & is.na(input_endYear) ~ "",
  TRUE ~ paste(input_startYear, input_endYear, sep = "/")))
``` 
 
### source

```{r}
distribution <- build_source_citation(distribution, checklists)
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("input_"))
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
```


# Description extension

## Pre-processing

```{r}
description <- input_descriptions
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

## Term mapping

### taxonID

```{r}
description %<>% mutate(taxonID = paste0("https://www.gbif.org/species/", input_verificationKey)) 
```

### description

```{r}
description %<>% mutate(description = input_description)
```

### type

```{r}
description %<>% mutate(type = input_type)
```

### language

```{r}
description %<>% mutate(language = "en")
```

### source

```{r}
description <- build_source_citation(description, checklists)
```

## Post-processing

Remove the original columns:

```{r}
description %<>% select(-starts_with("input_"))
```

Preview data:

```{r}
description %>% head()
```

Save to CSV:

```{r}
write_csv(description, here("data", "processed", "description.csv"), na = "")
```

# Create species profile extension

In this extension will express broad habitat characteristics of the species.

```{r}
species_profile <- input_speciesprofiles
```

## Term mapping

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(taxonID = paste0("https://www.gbif.org/species/", input_verificationKey))
```

### isMarine

```{r}
species_profile %<>% mutate(isMarine = input_marine)
```

### isFreshwater

```{r}
species_profile %<>% mutate(isFreshwater = input_freshwater)
```

### isTerrestrial

```{r}
species_profile %<>% mutate(isTerrestrial = input_terrestrial)
```

### source

```{r}
species_profile <- build_source_citation(species_profile, checklists)
```

## Post-processing

Remove the original columns:

```{r}
species_profile %<>% select(-starts_with("input_"))
```

Sort on `taxonID`:

```{r}
species_profile %<>% arrange(taxonID)
```

Preview data:

```{r}
species_profile %>% head()
```

Save to CSV:

```{r}
write.csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
```





