---
title: "Get taxa from checklists"
author:
- Peter Desmet
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how to retrieve the taxa from our checklists and create a verification spreadsheet.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Install packages:

```{r install_packages}
devtools::install_github("trias-project/trias@add_gbif_verify_synonyms")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(readr)
library(magrittr)


# Other packages
library(rlang)
library(lazyeval)
library(rgbif)
library(trias)
```

Set file paths:

```{r set_output_file_path}
taxa_output_file <- "../data/output/checklist_taxa.tsv"
checklist_metadata_file <- "../data/output/checklist_metadata.tsv"
```

# Select checklist datasets

These are the alien species checklist datasets we want to use and combine, ranked by trust (by experts):

```{r define_checklists}
checklists <- c(
  "9ff7d317-609b-4c08-bd86-3bc404b77c42", # Manual of the Alien Plants of Belgium
  "289244ee-e1c1-49aa-b2d7-d379391ce265", # Inventory of alien macroinvertebrates in Flanders, Belgium
  "98940a79-2bf1-46e6-afd6-ba2e85a26f9f"  # Checklist of non-native freshwater fishes in Flanders, Belgium
  # "b043c480-dd36-4f4f-aa82-e188753ff09d", # Catalogue of the Rust Fungi of Belgium
  # "3f5e930b-52a5-461d-87ec-26ecd66f14a3" # RINSE - Registry of non-native species in the Two Seas region countries (Great Britain, France, Belgium and the Netherlands)
)
```

Get metadata of these checklists from GBIF:

```{r get_checklist_metadata_from_gbif}
checklists_metadata <- map(checklists, function(x) rgbif::datasets(uuid = x))
checklists_metadata <- map_dfr(checklists_metadata, function(x) list(
  datasetKey = x$data$key,
  title = x$data$title,
  modified = x$data$modified,
  publisher = x$data$publishingOrganizationKey,
  doi = x$data$doi)
)
```

Preview metadata to ensure we have the correct checklists:

```{r preview_metadata}
checklists_metadata
```

We save the checklist details and order.

```{r save_checklist_info}
write_tsv(checklists_metadata, path = checklist_metadata_file)
```


# Get taxa

For each checklist, get the taxa from GBIF.

**Note**: here we return checklist taxa, not backbone taxa.

```{r get_taxa_from_gbif}
checklist_taxa_all_fields <- rgbif::name_lookup(
  datasetKey = checklists_metadata$datasetKey,
  origin = "source",# Only source taxa, not denormed higher classification taxa
  limit = 99999,
  return = "data")
```

Preview data:

```{r preview_taxa}
head(checklist_taxa_all_fields, 20)
```

Create a data frame with the information we want for each taxon, sorted by taxon key:

```{r select_columns}
checklist_taxa <- checklist_taxa_all_fields %>%
  select(key, scientificName, datasetKey, nubKey) %>%
  arrange(key)
```

Show some stats:

```{r show_taxa_stats_per_checklist}
checklist_taxa %>%
  group_by(datasetKey) %>%
  summarize(
    taxa = n(),
    taxa_not_in_backbone = sum(is.na(nubKey))
  )
```

Total number of taxa:

```{r show_total_taxa}
nrow(checklist_taxa)
```

# Filter on distribution

Our checklists might contain more than alien species in Belgium. We therefore use the `trias::has_distribution()` function to filter on the [distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml) information.

We will filter on:

1. [invasion stage](http://gbif.github.io/parsers/apidocs/org/gbif/api/vocabulary/EstablishmentMeans.html): `establishmentMeans` equal to one of `INTRODUCED`,  `NATURALISED`, `INVASIVE`, `ASSISTED COLONISATION`
2. [presence status](https://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceStatus.html): `status` equal to one of `PRESENT`, `COMMMON`, `RARE`, `IRREGULAR`
3. presence in Belgium (`country` equal to `BE`)

**Note**: we filter on distribution information of our _checklist taxa_, not backbone taxa. That is because backbone taxa contain distribution information from our own checklists _and_ other checklists, which we don't want to consider here.

E.g. compare:

* Distributions for _checklist_ taxon [Eriocheir sinensis](https://api.gbif.org/v1/species/140563012/distributions) (`140563012`)
* Distributions for _backbone_ taxon [Eriocheir sinensis](https://api.gbif.org/v1/species/2225776/distributions) (`2225776`)

Get distribution for all taxa and append `TRUE`/`FALSE` result as extra column. This step can take long.

```{r check_distributions_on_gbif}
checklist_taxa <- checklist_taxa %>%
  rowwise() %>%
  do_(interp(~ mutate(as.data.frame(.), 
                      valid_distribution  = trias::gbif_has_distribution(
                        taxon_key = .data$key,
                        country = "BE",
                        establishmentMeans = c("INTRODUCED", 
                                               "NATURALISED", 
                                               "INVASIVE", 
                                               "ASSISTED COLONISATION"),
                        status = c("PRESENT", "COMMMON", "RARE", "IRREGULAR"))
                      ))) %>%
  ungroup()
```

Preview results:

```{r preview_has_distribution}
checklist_taxa %>%
  select(key, scientificName, valid_distribution) %>%
  head(20)
```

Show some stats:

```{r show_distribution_stats_per_checklist}
checklist_taxa %>%
  group_by(datasetKey) %>%
  summarize(
    taxa = n(),
    taxa_with_valid_distribution = sum(valid_distribution),
    taxa_without_valid_distribution = sum(!valid_distribution)
  )
```

Only keep taxa with valid distribution:

```{r keep_valid_distribution}
checklist_taxa <- checklist_taxa %>%
  filter(valid_distribution == TRUE) %>%
  select(-valid_distribution)
```

Total number of taxa with valid distributions:

```{r show_total_valid_distribution_taxa}
nrow(checklist_taxa)
```

# Get backbone info

To make decisions regarding synonyms and higher classification, we need to retrieve GBIF backbone information, if present:

```{r retrieve_taxonomic_info_backbone}
taxon <- checklist_taxa %>% filter(is.finite(nubKey))
nub_keys <- checklist_taxa %>% 
  filter(is.finite(nubKey)) %>% 
  distinct(nubKey)

taxon_backbone_info <-  nub_keys %>%
  rowwise() %>%
  do_(interp(~ rgbif::name_usage(key = .$nubKey,
                         return = "data"))) %>%
  ungroup()
```

We select taxonomic information we are interested to:

```{r select_cols_backbone}
taxon_info <- c("key", "species", "genus", "family", "order",
                "class", "phylum", "kingdom",
                "rank", "speciesKey", "taxonomicStatus", 
                "acceptedKey", "accepted", "issues")

taxon_backbone_info <- taxon_backbone_info %>%
  select(one_of(taxon_info))
```

In case some information misses, add empty column(s):

```{r}
cols_to_add <- taxon_info[!which(taxon_info %in% 
           colnames(taxon_backbone_info))]

taxon_backbone_info <- map(cols_to_add, 
                           function(x) taxon_backbone_info %>% 
                             mutate(!! sym(x) := NA)) %>% 
  reduce(full_join, .init = taxon_backbone_info)
```

Next, we unlist column `issues`:

```{r unlist_issues}
taxon_backbone_info <- taxon_backbone_info %>%
  group_by_at(vars(-issues)) %>%
  summarize(issues = str_c(issues, collapse = ",")) %>%
  ungroup() %>%
  mutate(issues = ifelse(
    issues != "NULL",
    issues,
    NA
  ))
```

GBIF backbone issues returned: if multiple issues are present, they are separated by comma

```{r preview issues}
taxon_backbone_info %>% distinct(issues)
```

For improving readibility we rename the column `accepted` to `acceptedName` and add prefix `backbone_` or `checklist_` to all columns containing information from GBIF backbone or checklist respectively:

```{r add_prefix_backbone}
taxon_backbone_info <- taxon_backbone_info %>% rename(acceptedName = accepted)
colnames(taxon_backbone_info) <- paste0("backbone_", 
                                        colnames(taxon_backbone_info))
colnames(checklist_taxa) <- paste0("checklist_", 
                                        colnames(checklist_taxa))
```

We merge the information from checklists and GBIF backbone:

```{r}
combined_taxon_info <- checklist_taxa %>%
  left_join(taxon_backbone_info, 
            by = c("checklist_nubKey" = "backbone_key"))
```
```

Preview:

```{r preview_merged_final}
combined_taxon_info %>% head()
```

# Save output

Save the taxonomic information to a tsv file:

```{r}
write_tsv(combined_taxon_info, path = taxa_output_file, na = "")
```
