# Unify taxa

In this chapter we unify taxa on their `verificationKey`.

## Read data

1. Read taxa from [`data/interim/taxa_with_verification.csv`](https://github.com/trias-project/unified-checklist/blob/master/data/interim/taxa_with_verification.csv).

```{r unify_taxa-1}
taxa <- read_csv(here("data", "interim", "taxa_with_verification.csv"))
```

## Unify taxa

1. Separate multiple `verificationKey`s (if any) for single taxa.

```{r unify_taxa-2}
taxa <-
  taxa %>%
  
  # Separate multiple verification keys on pipe character in max 5 columns
  separate(
    verificationKey,
    into = c("verKey_1", "verKey_2", "verKey_3", "verKey_4", "verKey_5"),
    sep = "\\|",
    convert = TRUE,
    extra = "drop"
  ) %>%
  
  # Gather those 5 columns back to one (removing NA)
  gather(
    field,
    verificationKey,
    starts_with("verKey_"),
    na.rm = TRUE,
    convert = TRUE
  ) %>%
  
  # Drop field column
  select(-field)
```

2. Group by unique `verificationKey` to create unique/unified taxa:

```{r unify_taxa-3, echo = TRUE}
taxa_unify <-
  taxa %>%
  # Group by verificationKey across and within checklists
  group_by(verificationKey) %>%
  
  # Note contained checklists and taxa
  summarize(
    sourceChecklists = paste(sort(unique(datasetKey)), collapse = ","),
    sourceTaxa = paste(sort(unique(taxonKey)), collapse = ",")
  )
```

3. Extract `verificationKey` as a vector.

```{r unify_taxa-4}
verification_keys <-
  taxa_unify %>%
  pull(verificationKey)
```

4. Number of taxa: `r length(verification_keys)`

## Get GBIF backbone taxonomy information

Even though we stored some backbone information for most of our taxa in the previous steps, we want to start from scratch here and retrieve it from GBIF again, as 1) some taxon keys in `verificationKey`s will be new and 2) we want to store more attributes per taxon this time.

1. Get GBIF backbone taxonomy information.

```{r unify_taxa-get_backbone_info_from_gbif, cache = TRUE}
progress_bar <- progress_estimated(length(verification_keys))

backbone_info <-
  map_dfr(verification_keys, function(x) {
    progress_bar$tick()$print()
    rgbif::name_usage(
      key = x,
      return = "data"
    )
  })
```

2. Rename `accepted` to `acceptedName`.

```{r unify_taxa-5}
backbone_info <-
  backbone_info %>%
  rename(acceptedName = accepted)
```

3. Select columns of interest.

```{r unify_taxa-6}
backbone_info <-
  backbone_info %>%
  select(
    "key",
    "scientificName",
    "nameKey",
    "taxonID",
    "kingdom",
    "phylum",
    "class",
    "order",
    "family",
    "genus",
    "species",
    "datasetKey",
    "parentKey",
    "parent",
    "canonicalName",
    "authorship",
    "nameType",
    "rank",
    "taxonomicStatus",
    "acceptedKey",
    "acceptedName"
  )
```

4. Join backbone information with our unified taxa, so we keep `sourceChecklists` and `sourceTaxa`.

```{r unify_taxa-7}
taxa_unify <-
  taxa_unify %>%
  left_join(
    backbone_info, 
    by = c("verificationKey" = "key")
  )
```

5. Move columns `sourceChecklists` and `sourceTaxa` to the end.

```{r unify_taxa-8}
taxa_unify <-
  taxa_unify %>%
  select(-sourceChecklists,everything()) %>%
  select(-sourceTaxa,everything())
```

6. Preview merged information:

```{r unify_taxa-9}
taxa_unify %>% head()
```

7. Save to [CSV](https://github.com/trias-project/unified-checklist/blob/master/data/unified/taxa.csv).

```{r unify_taxa-11}
write_csv(taxa_unify, here("data", "unified", "taxa.csv"), na = "")
```
