---
title: "Verify synonyms"
author:
- Peter Desmet
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
# Other packages
library(rgbif)
library(trias)
library(googlesheets)
```

Set file path:

```{r input_file_path}
checklist_metadata_file <- "../data/output/checklist_metadata.tsv"
```


# Load data

Load checklist metadata (ranked by trust):

```{r load_checklist_metadata}
checklist_metadata <- read_tsv(checklist_metadata_file)
checklist_metadata
```

Load taxa information as previously wrangled via checklists and GBIF backbone:

```{r load_taxa}
taxa_file <- "../data/output/checklist_taxa.tsv"
taxa <- read_tsv(taxa_file)
```

Preview:

```{r}
taxa %>% head(n = 10)
```

Load verified taxa from spreadsheet in TrIAS Google Drive:

```{r load_googlesheet, eval = FALSE}
my_sheets <- googlesheets::gs_read(ss = "verification_spreadsheet",
                                   ws = "Sheet1")
```

Or via text file:

```{r load_text_file}
verified_synonyms_file <- "../data/output/verified_synonyms.tsv"
verified_synonyms <- read_tsv(verified_synonyms_file, 
                              col_types = cols(
                                backbone_taxonKey = col_integer(),
                                backbone_acceptedKey = col_integer(),
                                date_added = col_date())
                              )
```

Preview:

```{r}
verified_synonyms %>% head(n = 10)
```

# Assign verified key(s) to taxa

We need to verify synonyms and solve unmatches to GBIF backbone. Experts should verify taxa if one of the following conditions holds true:

1. no match with GBIF Backbone (`backbone_taxonKey` is not present)
2. taxonomic status not available or a synonym (`backbone_taxonomicStatus` not `ACCEPTED` or `DOUBTFUL`, see [GBIF taxonomic status page](http://rs.gbif.org/vocabulary/gbif/taxonomic_status.xml) for more details)

In all other cases, we can accept the `backbone_taxonKey` as provided by GBIF backbone:

```{r set_verified_key_equal_to_nubKey_or_NULL}
taxa <- taxa %>% 
  mutate(verification_key = if_else(is.finite(backbone_taxonKey) & 
                                  backbone_taxonomicStatus %in% 
                                  c("ACCEPTED", "DOUBTFUL"),
                                as.character(.data$backbone_taxonKey),
                                NULL))
```

A summary:

```{r stats}
taxa %>% count(with_verification_key = !is.na(verification_key))
```


We use function `verify_taxa()` from  `trias` package in order to 

1. retrieve verified keys which have been already assigned (stored in `verified_synonyms`)
2. update `verified_synonyms`. Updates include:
  - Updates from GBIF in the scientific names (`checklist_scientificName`, `backbone_scientificName` or `backbone_acceptedName`): those mostly consider slight spelling variations and will be updated in `verified_synonyms` authomatically (a message is returned).
  - New synonyms or no matches to GBIF backbone will be added to `verified_synonyms`.

The function reports also unused rows in `verified_synonyms`, but they will not be removed or altered.

```{r apply_verify_taxa}
verify <- trias::verify_taxa(taxa = taxa %>%
                               filter(is.na(verification_key)), 
                   verified_taxa = verified_synonyms)
```

## Save verification table

The data.frame `verify$verified_taxa` can be saved as text file and manually transferred to online spreadsheet:

```{r save_verification_df_to_tsv}
write_tsv(verify$verified_taxa, 
          path = "../data/output/verified_synonyms.tsv",
          na = "")
```

## New synonyms

The following synonyms are new and should be verified:

```{r new_synonyms}
verify$new_synonyms
```

## New unmatched taxa

The following taxa are not matched to GBIF backbone:

```{r new_unmatched_taxa}
verify$new_unmatched_taxa
```

## Updated names

Consequences of updates in GBIF backbone with no consequences in synonym relation.

### Updated backbone scientific names 

These backbone scientific names have been updated (but synonym relation stayed the same):

```{r updated_scientific_names}
verify$updated_scientificName
```

### Updated backbone accepted scientific names 

These backbone accepted names have been updated (but synonym relation stayed the same):

```{r updated_scientific_names}
verify$updated_acceptedName
```

### Updated issues

The backbone issues related to these taxa have been udpated:

```{r update_backbone_issues}
verify$updated_backbone_issues
```

## Unused taxa

These synonym relations are no longer used:

```{r}
verify$unused_taxa
```

## Update taxa

We can now update taxa with verified info:

```{r update_taxa}
taxa <- taxa %>% left_join(verify$verified_taxa %>%
                             separate_rows(checklists, sep = ",") %>%
                             rename(checklist_datasetKey = checklists) %>%
                             select(- c(date_added, remarks)))
```

Shows some stats:

```{r}
print(paste("Number of accepted taxa: ", 
            taxa %>% 
              filter(backbone_taxonomicStatus == "ACCEPTED") %>%
              nrow()))
print(paste("Number of doubtful taxa (no expert check needed): ", 
            taxa %>% 
              filter(backbone_taxonomicStatus == "DOUBTFUL") %>%
              nrow()))
print(paste("Number of taxa verified by expert: ",
            taxa %>% 
              filter(!is.na(verification_key)) %>%
              filter(! backbone_taxonomicStatus %in% c("ACCEPTED", "DOUBTFUL") |
                       is.na(backbone_taxonKey)) %>%
              nrow()))
print(paste("Number of taxa still to verify: ",
            taxa %>% 
              filter(is.na(verification_key)) %>%
              nrow())) 
```

Saved the updated taxa:

```{r save_udpated_taxa}
# save to csv -> taxa_by_nubKey_verified
taxa_verified <- write_tsv(taxa, 
                           path = "../data/output/verified_taxa.tsv",
                           na = "")
```
