---
title: "Verify synonyms"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    df_print: kable
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

Load taxa:

```{r}
# load from csv(taxa_by_nubKey)
```

Preview:

```{r}
# nubKey
# key
# dataset_scientificName
# datasetKey
# scientificName
# kingdom
# taxonomicStatus
# acceptedKey
# accepted
```

Load verified synonyms:

```{r}
# load from tsv(verified_synonyms)
```

Preview:

```{r}
# columns:
# nubKey
# scientificName
# acceptedKey
# accepted
# kingdom
# date_added
# verified_key
# remarks
```

## Find new synonym combinations

For every key, acceptedKey combination, where acceptedKey is not NA, update verified synonyms:

```{r}
# For every key, acceptedKey -> check in file

# If in file:
# If scientificname + acceptedScientificName different -> update + store in updated_names
# If same -> do nothing

# If not in file -> add new line + store in new_synonyms

# If in synonyms file but not found in grouped_taxa -> add " OUTDATED" to remarks + store in outdated_synonyms
```

## Results

These names have been updated (but synonym relation stayed the same):

```{r}
# updated_names
```

These synonym relations are no longer used:

```{r}
# outdated_synonyms
```

These synonym relations were added and should be verified:

```{r}
# new_synonyms
```

## Update taxa

Add `verified_key`

```{r}
# add column verified_key (is empty)
```

Accept all ACCEPTED and DOUBTFUL:

```{r}
# for taxonomicStatus = ACCEPTED or DOUBTFUL
# verified_key = nubKey
```

Update taxa_by_nubKey with verified synonyms info:

```{r}
# for taxonomicStatus = SYNONYM
# Look up nubKey + acceptedKey in verifiedSynonyms
# populate verified_key
```

Shows some stats:

```{r}
# Number of accepted taxa
# Number of doubtful taxa
# Number of verified synonyms
# Number of unverified synonyms
```

```{r}
# save to csv -> taxa_by_nubKey_verified
```
